---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# msp-ms

<!-- badges: start -->
<!-- badges: end -->

The goal of msp-ms is provide a concise code-base for the normalization and data processing required for the msp-ms method. 


Additionally, we provide a graphical user interface powered by shiny apps that allows for a user to utilize the method without requiring any R coding knowledge. 

So how does the msps data normalization process work?

1. Takes two input files from PEAKS and combines them
3. Normalizes values and then does a reverse log2 transformation
4. Looks for outliers across replicates. Removes them. 
5. Imputes data for missing values (likely to be very low intesity).
6. Figures out the locations of the detected clevages within the library of peptide sequences used. Is it cleaved at the N or C terminus, or both?
7. Calculates the fold change and the p/q value across experimental conditions using T-tests. 


## Combining Peaks file outputs

The files coming from peaks should have..... **Describe how to do this here**.



Importantly, this. data contains... what????


```{r}
### Loading the files ###
lfq_filename = "legacy/protein-peptides-lfq.csv"
#file "protein-peptides.csv" exported from PEAKS identification
id_filename = "legacy/protein-peptides-id.csv"

# Prepare the data for normalyzer analysis
prepared_data = prepare_for_normalyzer(lfq_filename,id_filename)
```




## Normalizing values.

msp-ms uses the .... package to do normalization under the hood. 


First we need to extract a design matrix. 

Then we can normalyze the data. 
```{r}
# Extracting design matrix
design_matrix = extract_design_matrix(prepared_data)

# normalyze the data 

source("R/normalyze.R")

normalyzed_data = normalyze(prepared_data,design_matrix)
```


## Handling Outliers

Here we use a dixon test from the outliers package to detect outliers from each of our replicates. 

```{r}
outliers = handle_outliers(normalyzed_data)
```



## Imputation of data 

We have a lot of missing, or 0 values. For these, we need to impute them so we can do downstream statistics
Data is imputated by ....

```{r}
imputed = impute(outliers)
```


## Joining with Library

Next we need to join everything with the sequences of the peptide library
```{r}
joined_with_library = join_with_library(imputed)
```



## Calcuating clevages
Next, we need to determine where the peptide sequences are cleaved. 

We check both the N and C terminus. 

Sequences are presented as the 4 amino acids on both sides of a cleavage. X indicates that there was nothing on that side in the library because it was cleaved close to the edge. 

```{r}
final_data = add_cleavages(joined_with_library)
```


## Conducting stats

Right now this is left to the user to do on their own. Will update after talkng to Bri.





